---
title: "in class_assign3"
output: html_document
name: Elvira Kalmurzaeva
Date: September 4, 2020
---

SUCCESS OF LEADER ASSASSINATION AS A NATURAL EXPERIMENT

One longstanding debate in the study of international relations concerns the
question of whether individual political leaders can make a difference. Some emphasize
that leaders with different ideologies and personalities can significantly affect the
course of a nation. Others argue that political leaders are severely constrained by
historical and institutional forces. Did individuals like Hitler, Mao, Roosevelt, and
Churchill make a big difference? The difficulty of empirically testing these arguments
stems from the fact that the change of leadership is not random and there are many
confounding factors to be adjusted for. 
In this exercise, we consider a natural experiment in which the success or failure
of assassination attempts is assumed to be essentially random.7 Each observation of
the CSV data set leaders.csv contains information about an assassination attempt.
Table 2.8 presents the names and descriptions of variables in this leader assassination data set. The polity variable represents the so-called polity score from the Polity Project. The Polity Project systematically documents and quantifies the regime types of all countries in the world from 1800. The polity score is a 21-point scale ranging from −10 (hereditary monarchy) to 10 (consolidated democracy). The result variable is a 10-category factor variable describing the result of each assassination attempt.

1. How many assassination attempts are recorded in the data? How many countries
experience at least one leader assassination attempt? (The unique() function, What is the average number of such attempts (per year) among these countries?

```{r}
leaders <- read.csv ("leaders.csv")
nrow(leaders)
#There are 250 assassination attempts recorded in the data. 
length(unique(leaders$country)) 
#88 countries experienced at least one leader assassination attempt.   
mean(tapply(leaders$country, leaders$year, length))
#The average number of leader assassination attempts is 2.45 among these countries. 
```

2. Create a new binary variable named success that is equal to 1 if a leader dies
from the attack and 0 if the leader survives. Store this new variable as part of the
original data frame. What is the overall success rate of leader assassination? Does
the result speak to the validity of the assumption that the success of assassination
attempts is randomly determined?

```{r}
leaders$result2 <- factor(leaders$result)
lev <- levels(leaders$result2)
lev
leaders$outcome <- as.integer (leaders$result2)
leaders$success <- ifelse(leaders$outcome %in% c(1,2,3,4),1,0)
mean(leaders$success)                       
#Overall leader assassinations attempts were successful for 21.6% It is too early to validate or disapprove the assumption of the the success of assassination attempts being randomly determined.
```

3. Investigate whether the average polity score over three years prior to an assassination
attempt differs on average between successful and failed attempts. Also,
examine whether there is any difference in the age of targeted leaders between
successful and failed attempts. Briefly interpret the results in light of the validity
of the aforementioned assumption.

```{r}
scoreSuccess <- mean(leaders[leaders$success == 1, "politybefore"])
scoreFail <- mean(leaders[leaders$success == 0, "politybefore"])
ageSuccess <- mean(leaders[leaders$success == 1, "age"])
ageFail <- mean(leaders[leaders$success == 0, "age"])
#The average difference in polity before between successful and failed attempts is one point, closer to democratic governance system. Which means that leaders of countries closer to democratic regimes are more likely to be killed. 
#The average difference in age of leaders is around four years, which may mean that the leaders who are older is more likely to die. There are two assumptions why this may happen: 1) due to the age their health conditions may be worse and they were not able to recover after the assassination attempt and 2)they may serve longer in the office which means that they may be targeted several times more than those who are younger. The several assassination attempts also increases chance of leaders to die. 
```

4. Repeat the same analysis as in the previous question, but this time using the
country’s experience of civil and international war. Create a new binary variable
in the data frame called warbefore. Code the variable such that it is equal to 1
if a country is in either civil or international war during the three years prior to
an assassination attempt. Provide a brief interpretation of the result.

```{r}
leaders$warbefore = ifelse((leaders$interwarbefore == 1 | leaders$civilwarbefore == 1), 1, 0)
SuccessWar <- mean(leaders[leaders$success == 1, "warbefore"])
FailWar <- mean(leaders[leaders$success == 0, "warbefore"])
#The overall rate of wars prior the successful assassination attempt is 35%, while before the  failed attempt is 37%. There is no great difference between attempts, therefore, we may assume that having a war before assassination does not really affect the outcome of the assassination attempt. However, the comparison of war before the assassination to countries years when there were no any attempts may tell us whether failed domestic politics (i.e. state of war is a proof of failed domestic politics) led to the assassination of leaders or did not effect at all. 

```

5. Does successful leader assassination cause democratization? Does successful
leader assassination lead countries to war? When analyzing these data, be sure
to state your assumptions and provide a brief interpretation of the results.

```{r}
scoreSuccessAfter <- mean(leaders[leaders$success == 1, "polityafter"])
scoreFailAfter <- mean(leaders[leaders$success == 0, "polityafter"])
#The overall rate of regime type demonstrates that after successful assassination of leaders countries were closer to democratic spectrum (-0.76), in comparison to failed attempts (-1.9). We may assume that failed assassinations led to toughening internal politics and introducing harsh measures to prevent any threat to the existing regime. 
leaders$warafter = ifelse((leaders$interwarafter == 1 | leaders$civilwarafter == 1), 1, 0)
SuccessWarAfter <- mean(leaders[leaders$success == 1, "warafter"])
FailWarAfter <- mean(leaders[leaders$success == 0, "warafter"])
#Moreover, the overall rate of international and civil wars after assassination attempts differs based on the outcome of the assassinations. About 20% of countries with successful attempts went to war, while 29.6% of countries with failed attempts went to war after. 
#We may assume that after failed attempts opposition or opposing leaders did not surrender and used violence causing civil war or international war. Failed assassination may lead them using another last resort. 
#In general, the success of the assassination attempts depends on various factors such as health condition of the leader, how badly she/he was injured, the quality of medical care during that period, but factors such as prior war experience, the regime type, age have various interpretations. Thus, not having the former set of factors at hand it is hard to state to what extend the results of the attempt are random. 
```

